---
title: "R tutorial: Choose the Most Appropriate Test"
author: "Leary Ortho Biostats Lab"
date: "10/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial will help you to determine the best test for your data. The example data set we are using here can be downloaded at [Box](https://missouri.box.com/s/70v91qllsnax3fzwyci2gam30llnu7rh) or [Github](https://raw.githubusercontent.com/zhengyes/rt/master/rExampleData.csv). The data are continous and describe values of 6 kinds of biomarkers for 144 animals by treatment.

### Flowchart

Here is the flowcahrt of choosing test.

```{r,echo=FALSE}
library(DiagrammeR)
mermaid('
graph TB
  A[Continuous Data]-->B{Normally Distributed<br> or Skewed}
  B-->C[Normally Distributed]
  B-->H[Skewed]
  C-->|2 groups| D{Paired<br>or Unpaired}
  C-->|2+ groups| G[\"ANOVA (post hoc)\"]
  D-->|unpaired| E[Two sample t test]
  D-->|paired| F[Paired t test]
  H-->|2 groups| I{Paired<br> or Unpaired}
  I-->|unpaired| J[Wilcoxon Rank Sum]
  I-->|paired| K[Wilcoxon Signed Rank]
  H-->|2+ groups|L[\"Kruskal-Wallis (post hoc)\"]
  
  style G fill:#c1f5f4
  style E fill:#c1f5f4
  style F fill:#c1f5f4
  style J fill:#c1f5f4
  style K fill:#c1f5f4
  style L fill:#c1f5f4
')
```

### What is hypothesis
Statistical hypothesis is an assumption about a population parameter. We can check whether or not a sample data is consistent with the statistical hypothesis, if not, the hypothesis is rejected.

There are two types of hypotheses. Null hypothesis H0 (which we try to reject with sample data) and alternative hypothesis Ha (opposite of H0). 

We need the p-value to determine whether or not we can reject H0. P-value is the probability of finding the observed, or more extreme, results when H0 is true. ("extreme" depends on how the hypothesis is being tested) 

If the test p-value is less than the significant level (usually 0.05) then we reject the H0, otherwise, we cannot reject H0.

### How to import data

After downloading the data to your computer, you need to find the path of the data file and import the data to R.

1. You can type the following code in console, then a "select file" window will pop up. Find the file you want to import and click "Open".
```{r eval=FALSE}
file.choose()
```

2. Now you know the path of the file, you can use read.csv function to read the .csv file into R. Type the following code in console but replace path between quotations with your file path. The new data set will be named "biomarker".
```{r}
biomarker = read.csv("C:\\Users\\zs7hm\\Desktop\\rExampleData.csv")
```

### Normality check
Many statistical tests require data to be normally distributed, this can be tested using a normality test, for example, Shapiro test. 
The null hypothesis of normality test is that there is no significant departure from normality. When the p is greater than .05, it fails to reject the null hypothesis and thus the assumption holds. Parametric methods (two sample t test, paired t test, ANOVA) need normality check, while the non-parametric methods (Wilcoxon rank sum, signed rank, Kruskal) can be used when data failed normality check.

Here is an example. We are going to use Shapiro test to test if biomarker MMP3 with treatment 0 is normally distributed. The null hypothesis of Shapiro test H0: data are normally distributed.

```{r, eval=FALSE}
# "which(biomarker$Treatment==0)" choose all subjects with treatment = 0, "MMP3" choose the MMP3 column.
shapiro.test(biomarker[which(biomarker$Treatment==0),"MMP3"])
```
The p-value = 0.3694, which is above 0.05. The following result suggests MMP3 with treatment 0 follows normal distrbution.
```{r, echo=FALSE}
shapiro.test(biomarker[which(biomarker$Treatment==0),"MMP3"])
```

### Paired or unpaired

#### paired

If you collect two measurements on the same experimental unit, then each pair of observations is closely related. In that case, you should use the paired t-test to test the mean difference between these dependent observations if your data is normally distributed and has same variance. If your data does not meet the normality or homogeneity of variance assumption, you should use nonparametric method Wilcoxon Signed Rank test.

#### Unpaired

If you randomly sample each set of items separately, under different conditions, the samples are independent. The measurements in one sample have no bearing on the measurements in the other sample, then the samples are unpaired. You should use the 2-sample t test to compare the difference in the means if your data is normally distributed and have equal variance. If your data does not meet the normality or homogeneity of variance assumption, use nonparametric method Wilcoxon Rank Sum test.

It is important to use the correct test to prevent wrong results. 

### Two sample t test
Two sample t test is used to determine if two population means $\mu_1$ and $\mu_2$ are equal. Null hypothesis H0: $\mu_1$ = $\mu_2$, alternative hypothesis Ha: $\mu_1$ $\neq$ $\mu_2$. We cannot reject the null hypothesis if p-value is greater than 0.05. Two sample t test is a parametric method, which means that the assumptions must be satisfied before applying it. If assumptions cannot be met, non-parametric methods can be used.

1. Assumptions
    + Data are continous
    + Samples are simply random samples from their respective populations, which means that each individual in the population has an equal chance of being selected in the sample. 
    + Two samples are independent
    + Data follow normal probability distribution
    + Variances of the two populations are equal

2. Goal: We want to compare the means of MMP3 with treatment 0 and treatment 5, to see if they are equal. We need to check assumptions first, if assumptions are met, we then apply two sample t test.

3. Code:

**Normality check:**
```{r,eval=FALSE}
# We need to check normality for both treatment groups. Treatment 0 group has been checked in Normality check section
shapiro.test(biomarker[biomarker$Treatment==5,"MMP3"])
```
The p-value of shapiro test is 0.09077, which is greater than 0.05. We cannot reject null hypothesis, so treatment 5 for MMP3 is normally distributed. We got the result in normally check section that treatment 0 for MMP3 is normally distributed. Therefore, both groups are normal.
```{r,echo=FALSE}
shapiro.test(biomarker[biomarker$Treatment==5,"MMP3"])
```
**Equal variance:**
```{r,eval=FALSE}
# Equal variance test
var.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==5,"MMP3"])
```
F test is used to check if two groups have the same variance. p-value = 0.2597, so we cannot reject the null hypothesis H0: true ratio of variances is 1. Equal variance asumption is satisfied.
```{r,echo=FALSE}
var.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==5,"MMP3"])
```
**Two sample t test:**
```{r,eval=FALSE}
# Two sample t test. (var.equal default is False, so you cannot ignore var.test = T) 
t.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==5,"MMP3"], var.equal = T)
```
Since all assumptions are met, we can use two sample t test now. p-value = 0.2594 > 0.05, so we cannot reject the null hypothesis H0: two populations have the same mean. Therefore, MMP3 treatment 0 and treatment 5 have the same mean. 
```{r,echo=F}
t.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==5,"MMP3"], var.equal = T)
```

4. Note:
Checking the assumptions is very important. If the normality check or the equal variance check fails, you need to use non-parametric methods Wilcoxon Rank Sum test, which will be introduced later. If only the equal variance check fails, you can use Welch's t test. You need to delete "var.equal = T" or change it to "var.equal = F" when you want to apply Welch's t test.

### Paired t test
Paired t test is used to determine whether the mean difference between two sets of observations is zero. The null hypothesis H0: $\mu_d$ = 0, alternative hypotheis Ha:$\mu_d$ $\neq$ 0. Null hypothesis is rejected when p-value less than 0.05. Paired t test is a parametric method, which means that the assumptions must be satisfied before applying it. If assumptions cannot be met, non-parametric methods can be used.

1. Assumptions
    + Data (differences for the matched-pairs) are continous
    + Data follow a normal probability distribution.
    + The sample of pairs is a simple random sample from its population. 
    
2. Goal:The null hypothesis for paired t test assumes that the **mean difference of two groups** equals to 0. We want to check whether or not the corresponding MMP3 and CTXII for each animal for treatment 0 have the same mean. 

3. Code:

**Normality check:**
```{r,eval=FALSE}
# Normality check for the difference 
shapiro.test(biomarker[biomarker$Treatment==0,"MMP3"] - biomarker[biomarker$Treatment==0,"CTXII"])
```
Shapiro test shows that the difference between two groups follows normal with p-value = 0.5969 > 0.05, so we can continue to use paired t test. 
```{r,echo=FALSE}
shapiro.test(biomarker[biomarker$Treatment==0,"MMP3"] - biomarker[biomarker$Treatment==0,"CTXII"])
```
**Paired t test**
```{r,eval=FALSE}
# Paired t test. (paired = F is default, you cannot ignore paired = T)
t.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==0,"CTXII"],paired = T)
```
The p-value = 2.2e-16, which is below 0.05. The conclusion is to reject the null hypothesis and that the means of MMP3 and CTXII are significantly different.
```{r,echo=FALSE}
t.test(biomarker[biomarker$Treatment==0,"MMP3"], biomarker[biomarker$Treatment==0,"CTXII"],paired = T)
```

4. Note:
Checking the assumptions is very important. If the normality check fails, you need to use non-parametric methods Wilcoxon Signed Rank test, which will be introduced later.

### ANOVA (post hoc)
ANOVA can make multiple comparison among more than two groups to see if all of them have the same mean. NUll hypotheis H0: $\mu_1$ = $\mu_2$ = $\mu_3$ =...= $\mu_k$, Ha: means are not all equal. The post hoc test can tell us which group has different population mean. 

1. Assumptions
    + Continuous data
    + Each group is normally distributed
    + Variances of populations are equal
    + Groups are independent
    + Each group is a simple random sample from its population
    
2. Goal: We want to compare means of three treatments (0,5 and 50) for MMP3.

3. Code:

**Normality check**
```{r,eval=FALSE}
# We need to check normality for all three treatment groups. Treatment 0 and 5 groups have been checked in Normality check and two sample t test sections.
shapiro.test(biomarker[biomarker$Treatment==50,"MMP3"])
```
The p-value of shapiro test is 0.4246, which is greater than 0.05. We cannot reject null hypothesis, so treatment 50 for MMP3 is normally distributed, and all three treatment groups are normally distributed.  
```{r,echo=FALSE}
shapiro.test(biomarker[biomarker$Treatment==50,"MMP3"])
```
**Equal variance**
```{r,eval=FALSE}
# Equal variance check. (You can still use var.test() which is F test, but you need to compare all the pairs. Using Bartlett test can make things easier because it makes multiple comparisons)
bartlett.test(MMP3~Treatment,data=biomarker)
```
The p-value = 0.4782 > 0.05, so we cannot reject the null hypothesis H0: three groups have equal variance. 
```{r,echo=FALSE}
bartlett.test(MMP3~Treatment,data=biomarker)
```
**ANOVA**
```{r,eval=FALSE}
# ANOVA (Treatment is just integer in the data set, we need to transfer it to factor)
a=aov(MMP3~as.factor(Treatment),data=biomarker)
summary(a)
```
You need to transfer treatment variable into factor. The p-value = 0.00607 < 0.05, so we can reject the null hypothesis H0: all three groups have the same mean and conclude that there is at least one group differ from others. 
```{r,echo=FALSE}
a=aov(MMP3~as.factor(Treatment),data=biomarker)
summary(a)
```
**Post hoc (Tukey)**
```{r,eval=FALSE}
# Post hoc analysis. Tukey is just one of the post hoc procedures which is commonly used.
TukeyHSD(a)
```
The p-value for 50-5 comparison is 0.0047412 < 0.05, the other two comparisons have p-value greater than 0.05. So, the Tukey test reveals that group 50 and 5 are significantly different. Because the lower bound and upper bound for 95% confidence interval are both greater than 0 (lwr = 5.260493, upr = 35.086174), we can conclude that mean of treatment 50 group is greater than mean of treatment 5 group. We also conclude that there is no significant difference between neither treatment 5 and treatment 0, nor treatment 50 and treatment 0.
```{r,echo=FALSE}
TukeyHSD(a)
```

4. Note:
Checking the assumptions is very important. If the normality check or equal variance assumption fails, you need to use non-parametric methods Kruskal-Wallis test, which will be introduced later. In the code above, you need to transfer treatments to factor variables.

### Wilcoxon Rank Sum (MannWhitney U test)
Wilcoxon rank sum test is used to test whether two samples are likely to derive from the same population (i.e., that the two populations have the same shape). you can also interpret this test as comparing the medians between the two populations. The null hypothesis H0: two populations distributions are equal, Ha: two populations distributions are not equal.

1. Goal:This test is a nonparametric alternative to the equal-variance t-test for use when the assumption of normality is not valid. This test uses the ranks of the values rather than the values themselves. We want to see if there is a difference between MMP1 and MMP2 for treatment 0.

2. Code:
```{r,eval=FALSE}
# Want to show normality is not satisfied
shapiro.test(biomarker[biomarker$Treatment==5,"MMP2"])
shapiro.test(biomarker[biomarker$Treatment==0,"MMP2"])
```
The Shapiro test shows that both groups are not normally distributed because their p-values are both lower than 0.05.
```{r,echo=FALSE}
shapiro.test(biomarker[biomarker$Treatment==5,"MMP2"])
shapiro.test(biomarker[biomarker$Treatment==0,"MMP2"])
```

```{r,eval=FALSE}
# Wilcoxon rank sum
wilcox.test(biomarker[biomarker$Treatment==0,"MMP2"],biomarker[biomarker$Treatment==5,"MMP2"])
```
The p-value is 3.937e-05, which is below 0.05. We conclude that MMP2 treatment 0 and 5 are significantly different.
```{r,echo=FALSE}
wilcox.test(biomarker[biomarker$Treatment==0,"MMP2"],biomarker[biomarker$Treatment==5,"MMP2"])
```

### Wilcoxon Signed Rank

This nonparametric test makes use of the sign and the magnitude of the rank of the differences (paired differences minus the hypothesized difference). It is one of the most commonly used nonparametric alternatives to the paired t-test. H0: difference between the pairs follows a symmetric distribution around zero. Ha: difference between the pairs does not follow a symmetric distribution around zero.

1. Goal: We want to compare corresponding MMP1 and MMP2 values for treatment 0. 

2. Code:
```{r,eval=FALSE}
# Want to show normality is not satisfied
shapiro.test(biomarker[biomarker$Treatment==0,"MMP1"])
shapiro.test(biomarker[biomarker$Treatment==0,"MMP2"])
```
Shapiro test p-value for MMP2 is 2.326e-05 < 0.05, so MMP2 is not normally distributed. 
```{r,echo=FALSE}
shapiro.test(biomarker[biomarker$Treatment==0,"MMP1"])
shapiro.test(biomarker[biomarker$Treatment==0,"MMP2"])
```

```{r,eval=FALSE}
# Wilcoxon signed rank
wilcox.test(biomarker[biomarker$Treatment==0,"MMP1"],biomarker[biomarker$Treatment==0,"MMP2"],paired = T)
```
The Wilcoxon Signed Rank test p-value = 7.105e-15, which is below 0.05. We conclude that MMP2 and MMP1 are significantly different for treatment 0.
```{r,echo=FALSE}
wilcox.test(biomarker[biomarker$Treatment==0,"MMP1"],biomarker[biomarker$Treatment==0,"MMP2"],paired = T)
```

### Kruskal-Wallis (post hoc)

Kruskal-Wallis test by rank is a non-parametric alternative to one-way ANOVA test, which extends the two-samples Wilcoxon test in the situation where there are more than two groups. It’s recommended when the assumptions of ANOVA test are not met. H0: medians of all groups are equal, Ha: at least one population median is different from at least one other group.

1. Goal: We want to compare three treatment groups of MMP2. The post hoc test can tell us which group is different.

2. Code:
```{r,eval=FALSE}
# Kruskal
kruskal.test(MMP2~as.factor(Treatment),data=biomarker)
```
The p-value for Kruskal-Wallis test = 0.0004431, which is below 0.05. We conclude that not all groups have the same median, at least one group has a different median. 
```{r,echo=FALSE}
kruskal.test(MMP2~as.factor(Treatment),data=biomarker)
```

```{r,eval=FALSE}
# Post hoc dunn test
library(FSA)
dunnTest(MMP2~as.factor(Treatment),data=biomarker,method = "bh")
```
The Dunn test p-values for 0-5 and 5-50 comparisons are less than 0.05, which indicates that treatment 0 and 5 have different medians, treatment 5 and 50 also have different medians.
```{r,echo=FALSE}
library(FSA)
dunnTest(MMP2~as.factor(Treatment),data=biomarker,method = "bh")
```
